Cross Validation Approaches for Penalized Cox Regression

### 2019-05-20 main.tex
Edits to Sections 4 and 5

\section{Introduction}
\par Since its original proposal \citep{Cox1972}, Cox proportional hazards regression has become the most common regression approach for analyzing survival data.  Cox regression utilizes a partial likelihood construction under an assumption of proportional hazards to estimate the regression coefficients without having to specify the underlying baseline hazard.  The ability to avoid choosing a specific parametric distribution for the survival time is very attractive, as time-to-event data are often poorly described by fully parametric models.

@@

The baseline hazard estimate we describe here differs from the one typically used in constructing deviance residuals \citep{Therneau1990}. Instead of the Nelson-Aalen estimator, deviance residuals are typically calculated based on a baseline hazard estimate that has been adjusted for the covariates in the model such as the one described in \citet{breslow1972}. In the context of cross-validation, however, this approach is problematic. Deviance residuals measure the difference between the fitted model and a saturated model; in Cox regression, this saturated model depends on the baseline hazard. Thus, a covariate-adjusted baseline hazard would mean that each fold is compared to a different saturated model. For this reason, we consider it important that the baseline hazard remains constant across folds when calculating cross-validated deviance residuals; this intuition is borne out by simulations involving various other approaches \note{supplement? otherwise, this sentence is a little tricky}\bnote{sure}.  Note that this is not the case for the Brier Score or Kullback Leibler Score (described in Section~\ref{Sec:accuracy}), which involve absolute loss functions, not comparisons with a saturated model.

@@

For this analysis, we fit elastic net models to the data in addition to lasso models, as these models often improve upon the lasso when many correlated genes are involved \citep{Zou2005}. The clinical variables stage, age and gender were included in the model without penalization, as it is reasonable to assume that all three have an effect on survival. In addition to the regularization parameter $\lam$, the elastic net model has an extra tuning parameter $\alpha$, where $\alpha=1$ corresponds to lasso and $\alpha=0$ corresponds to ridge regression. For $\alpha \in \{0.3, 0.5, 0.7, 0.9, 1\}$, we selected a regularization parameter $\lam$ that has the lowest cross-validation error for each of the four cross-validation methods described in Section~\ref{Sec:methods}.  Since the cross-validation metrics for each method are on different scales and thus not directly comparable, Table~\ref{tb:shedden} reports the cross-validated C-Index and Brier Scores for the CV-selected model at each value of $\alpha$.  As shown in the table, the elastic net model with $\alpha=0.3$ and $\lam$ selected by the cross-validated linear predictor approach has the best performance in terms of C Index and Brier Score.

@@

For each fixed $\alpha$, we compared the cross-validation errors (CVE) from the four cross-validation methods along the solution path. At $\alpha = 0.3$, CVEs are rescaled and plotted for all four methods along the same grid of $\lambda$ values in Figure~\ref{fig:CVE}. The selected $\lambda$ values, where a given CVE curve reaches its lowest point, are indicated with dotted lines.  The CVE curves for the standard and linear predictor approaches have greater curvature near their minima, and thus offer more clear support for the selected value of $\lam$.  In comparison, the deviance residual and V\&VH approaches are flatter near their minima, supporting a considerably wider range of $\lam$ values and leading to greater model uncertainty.   The selected $\lambda$ values, represented by the vertical dashed lines, and their corresponding models, reflected what is shown in simulation studies. The linear predictor approach, selecting smallest $\lambda$ values, was most liberal and selected 85 genes. While the standard approach and deviance residual approach selected 68 genes and 60 genes respectively, Verweij and Van Houwelingen's approach only selected 34 genes and performed most conservatively on this data set. Even though we only illustrated the models at $\alpha = 0.3$, same patterns are repeated when $\alpha$ is fixed at other values.

\begin{figure}[h]
  \centering
  \includegraphics[height= 8cm ]{./manuscript_figure/shedden.png}
  \caption{\label{fig:CVE}Rescaled Cross Validated Errors at $\alpha = 0.3$. The horizonal axis is $\log(\lambda)$. The vertical axis is the cross-validated error rescaled to a range from 0 to 1. After the rescaling, the Cross Validated Error of four different methods can be plotted together for comparison. The vertical dashed lines represent the $\lambda$ selected by a specific method. The linear predictor approach selected the smallest $\lambda$ and Verweij and Van Houwelingen's approach selected the greatest $\lambda$}
\end{figure}

We carried out further data analysis based on the elastic net model with $\lam = 0.650$ and $\alpha = 0.3$, selected by the cross-validated linear predictor approach. Of the 85 genes that were selected by this model, 78 were well-annotated.  We carried out an enrichment analysis of these 78 genes using ToppGene \citep{Chen2009c}; the five most significantly enriched molecular functions are listed in Table~\ref{Tab:go}.

\begin{table}[ht]
\centering
\caption{\label{Tab:go} Five most significantly enriched molecular functions among genes selected by the linear predictor method with the elastic net ($\alpha=0.3$).  The five terms have a false discovery rate of 3\% using the Benjamini-Hochberg method. \note{I'm not sure whether we should include Selected and Annotated or not...if we do, we'll have to describe them better.}}
\begin{tabular}{llrrr}
\toprule
ID & Name & p & Selected & Annotated\\
\midrule
GO:0008134 & transcription factor binding & 0.00015 & 10 & 584\\
GO:0035257 & nuclear hormone receptor binding & 0.00057 & 5 & 161\\
GO:0019869 & chloride channel inhibitor activity & 0.00061 & 2 & 9\\
GO:0003700 & DNA-binding transcription factor activity & 0.00068 & 14 & 1264\\
GO:0005102 & signaling receptor binding & 0.00083 & 16 & 1601\\
\bottomrule
\end{tabular}
\end{table}

Overall, the results are consistent with our understanding of cancer: many of the genes selected by the elastic net model are involved in signaling and the binding of transcription factors and receptors.  These are precisely the pathways that tend to be disrupted in cancer.  Chloride channels have also recently been implicated in several cancers as well as therapeutic targets \citep{Peretti2015}.

Of particular interest is the fact that three of the selected genes (PIK3R1, PDPK1, and ITGB1) are involved in the PTEN pathway.  The PTEN gene is an important tumor suppressor; mutations in PTEN are among the most common genetic changes found in human cancers and have been linked to cancers of the lung, breast, prostate, brain, and more.  Time-to-event analyses like this one of gene expression data provide valuable insights for developing gene signatures that can help to identify patients with more severe disruptions to important pathways such as PTEN.  These signatures thereby inform prognosis and impact treatment decisions.

@@

Conducting cross-validation is a crutial step in fitting penalized Cox regression model. In this paper, we proposed two intuitive ways to conduct cross-validation for Cox Regression: cross-validated linear predictor and cross-validated deviance residuals. We conducted a series of simulation experiments to study and compare their performance with two existing approaches, the standard cross-validation and Verweij and Van Houwelingen's cross-validated partial likelihood, in selecting tuning parameters when fitting penalized regression models.

Despite the fact Verweij and Van Houwelingen's cross-validated partial likelihood has been widely implemented, its predictive performance seems to suffer, particularly when the dimension of the data grows and signals become weak. As is illustrated in both simulation studies and real data example, Verweij and Van Houwelingen's approach tends to perform conversatively, selecting fewer genes than all other models. In the lung cancer data example, the number of genes selected by other methods almost doubled the number of genes selected by Verweij and Van Houwelingen's approach. From an application standpoint, using Verweij and Van Houwelingen's approach in high-dimensional datasets may lead to more false negatives and miss important true findings.

While the standard approach was the most liberal one across all simulation scenarios and yielded the best predictive performance, this approach limited by its stability issues which are carefully discussed in Section~\ref{Sec:stability}. The cross-validated linear predictor approach performs closely to the standard approach and demostrates an attractive balance of predictive performance and numerical stability.  We would recommend using the proposed linear predictor approach in practice for selecting regularization parameters in penalized regression.

>>>>

\section{Introduction}

\note{I hate to change the name again, but what do you think about calling it ``basic'' cross-validation instead of ``standard''?  My concern is that it's not really standard in the sense of ``the thing most people do''; I think most people use V\&VH since that's the default in glmnet.}

\par Since its original proposal \citep{Cox1972}, Cox proportional hazards regression has become the most common regression approach for analyzing survival data.  Cox regression utilizes a partial likelihood construction under an assumption of proportional hazards to estimate the regression coefficients without having to specify the underlying baseline hazard.  The ability to avoid choosing a specific parametric distribution for the survival time is very attractive, as time-to-event data are often poorly described by fully parametric models.

@@

The baseline hazard estimate we describe here differs from the one typically used in constructing deviance residuals \citep{Therneau1990}. Instead of the Nelson-Aalen estimator, deviance residuals are typically calculated based on a baseline hazard estimate that has been adjusted for the covariates in the model such as the one described in \citet{breslow1972}. In the context of cross-validation, however, this approach is problematic. Deviance residuals measure the difference between the fitted model and a saturated model; in Cox regression, this saturated model depends on the baseline hazard. Thus, a covariate-adjusted baseline hazard would mean that each fold is compared to a different saturated model. For this reason, it is important that the baseline hazard remains constant across folds when calculating cross-validated deviance residuals; this intuition is borne out by simulations involving various other approaches (see Supporting Information) \note{do we actually have a supporting information document written up?}.  Note that this is not the case for the Brier Score or Kullback Leibler Score (described in Section~\ref{Sec:accuracy}), which involve absolute loss functions, not comparisons with a saturated model.

@@

For this analysis, we fit elastic net models to the data in addition to lasso models, as these models often improve upon the lasso when many correlated genes are involved \citep{Zou2005}. The clinical variables stage, age and gender were included in the model without penalization, as it is reasonable to assume that all three have an effect on survival. In addition to the regularization parameter $\lam$, the elastic net model has an extra tuning parameter $\alpha$, where $\alpha=1$ corresponds to lasso and $\alpha=0$ corresponds to ridge regression. For $\alpha \in \{0.3, 0.5, 0.7, 0.9, 1\}$, we selected a regularization parameter $\lam$ that has the lowest cross-validation error for each of the four cross-validation methods described in Section~\ref{Sec:methods}.  Although each method can be used to select a pair of $\alpha$ and $\lam$ values, the methods are on different scales and cannot be directly compared with each other.  For this reason, Table~\ref{tb:shedden} reports the cross-validated C-Index and Brier Scores for each method's choice of $\lam$ at each value of $\alpha$.  As shown in the table, the elastic net model with $\lam$ and $\alpha=0.3$ selected by the cross-validated linear predictor approach has the best performance in terms of C Index and Brier Score.

@@

For each fixed $\alpha$, we compared the cross-validation errors (CVE) from the four cross-validation methods along the solution path. At $\alpha = 0.3$, CVEs are rescaled and plotted for all four methods along the same grid of $\lambda$ values in Figure~\ref{fig:CVE}. The selected $\lambda$ values, where a given CVE curve reaches its lowest point, are indicated with dotted lines.  The CVE curves for the standard and linear predictor approaches have greater curvature near their minima, and thus offer more clear support for the selected value of $\lam$.  In comparison, the deviance residual and V\&VH approaches are flatter near their minima, supporting a considerably wider range of $\lam$ values and leading to greater model uncertainty.   The selected $\lambda$ values, represented by the vertical dashed lines, and their corresponding models, reflect what was observed in the simulation studies. The linear predictor approach was most liberal and selected 85 genes. While the standard approach and deviance residual approach selected 68 genes and 60 genes respectively, Verweij and Van Houwelingen's approach was most conservative, selecting only 34 genes.  Similar patterns were observed for other $\alpha$ values.

\begin{figure}[h]
  \centering
  \includegraphics[height= 8cm ]{./manuscript_figure/shedden.png}
  \caption{\label{fig:CVE}Rescaled Cross Validated Errors at $\alpha = 0.3$. The horizonal axis is $\log(\lambda)$. The vertical axis is the cross-validated error rescaled to a range from 0 to 1 in order to allow plotting of the four methods on a common axis. The vertical dashed lines represent the $\lambda$ selected by a specific method. The linear predictor approach selected the smallest $\lambda$ and Verweij and Van Houwelingen's approach selected the greatest $\lambda$.}
\end{figure}

As mentioned earlier, for the elastic net model with tuning parameters chosen by cross-validated linear predictors ($\lam = 0.650$, $\alpha = 0.3$), 85 genes were selected, of which 78 were well-annotated.  We carried out an enrichment analysis of these 78 genes using ToppGene \citep{Chen2009c}; the five most significantly enriched molecular functions are listed in Table~\ref{Tab:go}.  Overall, the results are consistent with our understanding of cancer: many of the genes selected by the elastic net model are involved in signaling and the binding of transcription factors and receptors.  These are precisely the pathways that tend to be disrupted in cancer.  Chloride channels have also recently been implicated in several cancers as therapeutic targets \citep{Peretti2015}.

\begin{table}[ht]
\centering
\caption{\label{Tab:go} Five most significantly enriched molecular functions among genes selected by the linear predictor method with the elastic net ($\alpha=0.3$).  The five terms have a false discovery rate of 3\% using the Benjamini-Hochberg method. Enrichment: frequency of the annotation among the selected genes divided by its frequency in the genome as a whole.}
\begin{tabular}{llrrr}
\toprule
ID & Name & Selected & Enrichment & p \\
\midrule
GO:0008134 & transcription factor binding & 10 & 4.1 & 0.00015 \\
GO:0035257 & nuclear hormone receptor binding & 5 & 7.4 & 0.00057 \\
GO:0019869 & chloride channel inhibitor activity & 2 & 53.2 & 0.00061 \\
GO:0003700 & DNA-binding transcription factor activity & 14 & 2.6 & 0.00068 \\
GO:0005102 & signaling receptor binding & 16 & 2.4 & 0.00083 \\
\bottomrule
\end{tabular}
\end{table}

Of particular interest is the fact that three of the selected genes (PIK3R1, PDPK1, and ITGB1) are involved in the PTEN pathway.  The PTEN gene is an important tumor suppressor; mutations in PTEN are among the most common genetic changes found in human cancers and have been linked to cancers of the lung, breast, prostate, brain, and more.  Time-to-event analyses like this one of gene expression data provide valuable insights for developing gene signatures that can help to identify patients with more severe disruptions to important pathways such as PTEN, thereby informing prognosis and impacting treatment decisions.

@@

Cross-validation is a crutial step in fitting penalized Cox regression models. In this paper, we propose two cross-validation methods for Cox Regression: cross-validated linear predictors and cross-validated deviance residuals. We conducted a series of simulation experiments to study and compare their performance with two existing approaches, the standard cross-validation and Verweij and Van Houwelingen's cross-validated partial likelihood, in selecting tuning parameters when fitting penalized regression models.

Despite the fact Verweij and Van Houwelingen's cross-validated partial likelihood has been widely implemented, its predictive performance appears suboptimal, particularly when the dimension of the data is large and/or signals are weak. As illustrated in both simulation studies and real data example, Verweij and Van Houwelingen's approach tends to perform conservatively, selecting fewer features than all other models. In the lung cancer data example, the number of genes selected by other methods almost doubled the number of genes selected by Verweij and Van Houwelingen's approach. From an application standpoint, using Verweij and Van Houwelingen's approach in high-dimensional applications may lead to more false negatives and miss important true findings.

While the standard approach performed well in terms of predictive performance, this approach suffers from a lack of numerical stability as discussed in Section~\ref{Sec:stability}. The cross-validated linear predictor approach performed very similarly to the standard approach in terms of predictive performance, but without any problems due to numerical stability.  We recommend using the proposed linear predictor approach in practice for selecting regularization parameters in penalized Cox regression.
